Merging merges, more or less

### 2015-04-05 2015mergingmerges/section/01-Introduction.tex
Hide material in intro, and shift invariants to the end

Despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
Unlike the simple access patterns of append or zip, the merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.
Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.

Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.
However, for querying large data sets, these kind of combinators are essential.
The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.

In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.

Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.

@@

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.

Consider the following program.
Suppose we have two sources of integers from disk, wish to filter each and pair them together, then perform some action for each pair:

@@

In this paper, we present:
\begin{itemize}
\item a kind of deterministic finite automata, for expressing streaming computations with value-dependent rates (\S\ref{s:Machines});
\item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
\item invariants required of such machines (\S\ref{s:Machines:Invariants});
\item and an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging}).
\end{itemize}

>>>>

Despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
Unlike the simple access patterns of append or zip, the merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.

% -- Use the space to argue why merges are needed, not the other way around.
% -- This material would be better in the conclusion.

% Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
% At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.

% Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.

% For querying large data sets merge-like combinators are essential.
% The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
% Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
% Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.

% In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
% This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
% Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.

% Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.

\TODO{Insert key example demonstrating some function involving merges that needs to be fused.}
\TODO{The intro, including the following bulleted list should fit all on the first page.}

We make the following contributions:
\begin{itemize}
\item a kind of deterministic finite automata, for expressing streaming computations with value-dependent rates (\S\ref{s:Machines});
\item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
\item invariants required of such machines (\S\ref{s:Machines:Invariants});
\item and an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging}).
\end{itemize}

@@

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
Consider the following program.
Suppose we have two sources of integers from disk, wish to filter each and pair them together, then perform some action for each pair:

@@

### 2015-04-05 2015mergingmerges/section/02-Machines.tex
contributions
Hide material in intro, and shift invariants to the end

\subsection{Combinators}
As an example of @filter p xs@ combinator expressed as a machine, see figure~\ref{fig:com:filter}.
Here, the initial state is a @Pull xs@, an attempt to read from the input stream @xs@.
If the read is successful, we take the @Some xs@ transition leading to $@If@~(p~xs_e)$ - the predicate applied to the currently read element of @xs@.
If the predicate is true, we execute $@Out@~o~xs_e$, writing the current element of @xs@ to the output channel named @o@; otherwise, we release the element of @xs@ and go back to the @Pull@.
Again, this release is not strictly necessary for expressivity or code generation, but is essential for synchronisation during merging.
Finally, if the @Pull xs@ fails and there is no data left to read, we signal that the output stream has finished, and then finish the entire machine.

@@

\subsection{Invariants}
\label{s:Machines:Invariants}
Not all machines are valid or meaningful, and we wish to rule them out.
For example, a machine with an initial state of $@If@~(xs_e~>~5)~n$ is meaningless because there is no element of $xs$ that has been read from an input. 
On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but having explicit releases makes synchronising two machines easier.

The invariants are:
\begin{itemize}
\item each function mentioned in @Out@, @If@ or @Update@ can only refer to previously read values;
\item each successful @Pull@ must be @Released@ before another @Pull@ can be made;
\item all inputs must either be finished by an empty @Pull@ or @Close@d at @Done@;
\item all outputs must be finished with @OutDone@ at @Done@;
\item inputs cannot be read after an empty @Pull@ or after they are @Close@d;
\item outputs cannot be written after they are finished with @OutDone@.
\end{itemize}

To perform this invariant checking, we annotate each state with a set of available information at that state.
The DFA machines simplify the merging algorithm, but lose binding and scope information which would be implicit in a structured language.
We need to reconstruct and check the scope information, similar to a type state system\CITE.

The available event information can be that there exists some value that can be used, or that a channel is finished or closed.

\begin{tabbing}
MMMM \= M \= \kill
@Event@ \> $=$  \> $@Value@~n$ \\
       \> $~|$ \> $@Finished@~n$ \\
       \> $~|$ \> $@Closed@~n$ \\
\end{tabbing}

To begin, we annotate the initial state with an empty set of information: at the start of the machine, we have no read values available, nor is it known (or likely!) that any channels are finished.
Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions respectively.
If the output state has already been annotated, we check that the two annotations are the same, modulo values produced by @Out@.
The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
These @Out@ values are then implicitly released.

For each state, we check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, functions of @Out@ and @If@ require their values to be present, and so on.

\input{figures/Invariants.tex}

\subsubsection{Generation}
Figure~\ref{fig:inv:generation}.

Each transition modifies its input's annotation, depending on the state type and the transition label.
For example, the @Some@ output edge of a @Pull@ adds new information that a @Value@ is available.

$$ \CheckOut{\Gamma}{s}{a} $$
Under environment $\Gamma$ and machine $m$, the state $s$ has available information $a$.

$$ \CheckOutTrans{l}{a} $$
Under environment $\Gamma$ and machine $m$, the transition from state $s$ to $t$ under label $l$ has available information $a$.

$$ \CheckStateType{s}{T} $$
In machine $m$, state $s$ has state type $T$.

\subsubsection{Checking}
Figure~\ref{fig:inv:checking}.

We check each state type against its final annotation. For example, @Release@ requires that there exists a @Value@ to release.

$$ \StateOK $$
Under environment $\Gamma$ and machine $m$, the state $s$ has a valid annotation.

$$ \fvs{f} \subset a $$
Any free variables mentioned in a function must exist as @Value@s in the annotation, in order for the function to use these values.

\subsubsection{Environment closure}
Figure~\ref{fig:inv:closure}.

We must compute the transitive closure of each state's annotation.
Given two environments, the final environment and unvetted environment, states are pulled from the unvetted environment and checked against the final environment.
If the state already exists in the final environment and the annotations are equal, that state has been computed and can be removed from the unvetted environment.
If the state exists in both environments but they are different, either the only differences are output values, and we proceed with the intersection of the two, or if there are other differences, the machine is invalid.
Finally, if the state does not exist in the final environment, we add it to the final environment, compute all output transitions of the state, and add them to the unvetted environment.

\TODO{Perhaps this would be described better as a function than an inference rule.}
The problem, of course, is that state machines are not trees, so a derivation tree for typing does not make much sense.

$$ \EnvGrow{\Gamma}{p}{\Gamma'} $$
Under environment $\Gamma$ and machine $m$, with $p$ as an unvetted environment to be merged into $\Gamma$, the transitive closure of adding transitions is $\Gamma'$.

\subsubsection{Entire machine}
Figure~\ref{fig:inv:entire}.

First, the environment closure is computed, starting with the initial state of the machine having no available information.
Next, every state of the machine is checked to be valid against the environment closure.

$$ m~@ok@ $$
The entire machine is valid.

\subsection{Code extraction}

Extracting imperative code from these machines is quite straightforward.
Variables are created for each @Pull@ source, to store a single value read from the source, and each output channel has a variable to store its latest output and its current state.
Each state is converted to a basic block and given a unique identifier as its label.

For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the Figure~\ref{fig:extract:filter}.

\begin{figure}
\begin{code}
run :: IO Int -> (Maybe Int -> IO ()) -> IO ()
run pull_xs when_o = l10
 where
  l10 = do  x  <- pull_xs
            case x of
             Nothing -> l90
             Just x' -> l15 x'

  l15 x | p x       = l20 x
        | otherwise = l30 x

  l20 x =   when_o (Just x)  >> l30 x

  l30 x =   l10

  l90   =   when_o Nothing  >> l91

  l91   =   return ()
\end{code}
\caption{Extracted code for filter}
\label{fig:extract:filter}
\end{figure}

>>>>

\subsection{Combinators}
\label{s:Machines:Combinators}
As an example of @filter p xs@ combinator expressed as a machine, see figure~\ref{fig:com:filter}.
Here, the initial state is a @Pull xs@, an attempt to read from the input stream @xs@.
If the read is successful, we take the @Some xs@ transition leading to $@If@~(p~xs_e)$ - the predicate applied to the currently read element of @xs@.
If the predicate is true, we execute $@Out@~o~xs_e$, writing the current element of @xs@ to the output channel named @o@; otherwise, we release the element of @xs@ and go back to the @Pull@.
Again, this release is not strictly necessary for expressivity or code generation, but is essential for synchronisation during merging.
Finally, if the @Pull xs@ fails and there is no data left to read, we signal that the output stream has finished, and then finish the entire machine.

@@
